---
title: "Processing METAR weather reports"
author: "David Failing"
date: "2/17/2021"
output: 
  pdf_document:
    latex_engine: xelatex
---
Following along with the vignette athttps://cran.r-project.org/web/packages/pmetar/vignettes/pmetar.html.

# Downloading a current METAR weather report
For downloading a METAR report we need to know an airport four letters ICAO code, International Civil Aviation Organization, or three letters IATA code, International Air Transport Association.

Let’s download a current METAR weather report for Warsaw Okecie Airport. Its ICAO, International Civil Aviation Organization, code is EPWA. A report can be got from Aviation Weather Center https://www.aviationweather.gov/metar

```{r, message = FALSE, warning = FALSE, echo = FALSE}
library(dplyr)
library(pmetar)
```

```{r, message = FALSE, warning = FALSE}
# Let’s download a current METAR weather report for Warsaw Okecie Airport.
metar_get("EPWA")

# Now let’s take a look at Newark Liberty International Airport, EWR IATA code. 
# This type of code you find on your airplane tickets.
metar_get("EWR")
```

# Downloading historical METAR weather report

The function `metar_get_historical` allows to download METAR weather reports for a defined period of time. The default online source of METAR reports, from = “iastate” is the Iowa Environmental Mesonet web page of Iowa State University ASOS-AWOS-METAR http://mesonet.agron.iastate.edu/AWOS/.

```{r, message = TRUE, warning = TRUE}
dm <- metar_get_historical("JFK", start_date = "2020-06-25", end_date = "2020-06-29", from = "iastate")
head(dm)
```

The second backup source, from = “ogimet” is Weather Information Service provided by Ogimet http://www.ogimet.com/. Please take into consideration that Ogimet usually blocks too frequent requests for data due to the server overload, the requested period is limited to 31 days and for the most of airports METAR reports are available from the beginning of the year 2005.

**The message here is to save your data in .csv files once you download it, and not have to download it again, so that you don't get blocked for too frequent requests. `ogimet` is also limited to 31 day chunks at a time.**

```{r, message = TRUE, warning = TRUE}
metar_get_historical("JFK", start_date = "2020-06-25", end_date = "2020-06-29", from = "ogimet")
```

We will parse the last report from the above `dm.` In historical reports dates and hours are placed at the beginning of texts. Normally it’s extracted and parsed, but for now let’s remove it

```{r, message = TRUE, warning = TRUE}
(dm[length(dm)])
my_report <- substr(dm[length(dm)], 14, nchar(dm[length(dm)]))
my_report
```

It has to be noted that the report part after the **RMK**, remarks, is not analyzed, except the temperature information, which is provide more precisely in remarks.

**So it appears that this vignette does not deal with the part that I actually need to parse for dust.**

# Decoding main informatin from a single METAR weather report

The first element **METAR** indicates the text consist of a METAR weather report. If a report was issued under special circumstances, a text SPECI replaces METAR.

**This is good since it's still 5 characters at the start regardless.**

# Airport.
The second four letters element, **KJFK** identifies an airport from which a METAR issued. We can extract it

```{r, message = TRUE, warning = TRUE}
metar_airport(my_report) # Vectorized so you can feed multiple reports.
```
and find the **KJFK** geographical coordinates, elevation, airport IATA code, airport name and source of information.

```{r, message = TRUE, warning = TRUE}
metar_location(metar_airport(my_report))
```

# Day and time
The third element **282355Z** includes a day of a month, a time and a time zone.

```{r, message = TRUE, warning = TRUE}
metar_day(my_report)
metar_hour(my_report)
metar_time_zone(my_report)
```

The fourth element, in our case *AUTO* informs that a report was generated automatically. If a report was manually corrected it is COR. This element is not taken into consideration by the package pmetar.

# Wind speed and wind direction

Next, there is the text **07007KT** where three first digits informs about a wind direction in degrees. Two next digits are a wind speed and the letters the end define units, here KT, hence a wind speed is in knots.


```{r, message = TRUE, warning = TRUE}
metar_dir(my_report)
metar_speed(my_report, metric = TRUE)
metar_speed(my_report, metric = FALSE)
```

The function metar_speed reports a wind speed in m/s with the default value of the parameter `metric = TRUE`, or in knots when `metric = FALSE`.

When a wind direction varies, a METAR report has additional component, like *140V200*, which informs that a wind direction fluctuates from 140 to 200 degrees.

```{r, message = TRUE, warning = TRUE}
variable_direction_METAR <- "EPWA 281830Z 18009KT 140V200 9999 SCT037 03/M01 Q1008 NOSIG"
metar_dir(variable_direction_METAR)
```

In this case an output is character what can be useless for statistical calculations. If only main direction in the numeric format is needed, it is possible to set the parameter `numeric_only = TRUE.`

```{r, message = TRUE, warning = TRUE}
metar_dir(variable_direction_METAR, numeric_only = TRUE)
```

# Visibility

The part **10SM** is the visibility. In this case it’s 10 statue miles. With the default value of the paremeter `metric= TRUE` we get output in meters. For `metric = FALSE` in statute miles.

```{r, message = TRUE, warning = TRUE}
metar_visibility(my_report, metric = TRUE)
metar_visibility(my_report, metric = FALSE)
```

# Weather conditions

The function `metar_wx_codes` extracts and parses the below weather conditions codes:

```{r, message = TRUE, warning = TRUE}
metarWXcodes
```

In our METAR examples part **-RA** informs about weather conditions.

**Note, though, that this gives plaintext and not the codes list. Perhaps dig into the package specification to pull those easily**

```{r, message = TRUE, warning = TRUE}
metar_wx_codes(my_report) # No recorded weather

metar_wx_codes("202002022205 METAR KEWR 022205Z AUTO 24008KT 6SM -RA -SN BR SCT006 BKN014 OVC024 02/01 A2954 RMK T00200010 MADISHF")
```

# Cloud coverage

Next part, **SCT028 SCT035 BKN079**, informs about a cloud coverage

```{r, message = TRUE, warning = TRUE}
metar_cloud_coverage(my_report)
```

# Temperature and dew point

The temperature and the dew point can be extracted from two elements of a METAR report, before the RMK remarks marker *23/20* which can be found in the most reports. Or from the part after the RMK remarks marker *T02300200*, more precise but not always available.

The temperature is coded in Celsius degrees, here **23**/20, or more detailed in T0**230**0200.

```{r, message = TRUE, warning = TRUE}
metar_temp(my_report)
```

If there is a letter M in the front of two digits, M23/00, or there is a digit one after T, T1230, the temperature is below zero Celsius degrees.

The dew point can be decoded from the last two digits 23/**20**, or more detailed from T02300**200**. Dew points below zero Celsius degrees are decoded in the same method as above. For example 04/**M**03 or T0039**1**033 mean that the dew point temperature is -3 Celsius degrees or more precisely -3.3 Celsius degrees.

```{r, message = TRUE, warning = TRUE}
metar_dew_point(my_report)
```

Here there is a report with the more precise temperature information in the RMK remarks part.

```{r, message = TRUE, warning = TRUE}
metar_temp("202001010851 METAR KEWR 010851Z 27010KT 10SM FEW030 BKN070 BKN100 BKN210 04/M03 A2969 RMK SLP054 T00391033 52012")
metar_dew_point("202001010851 METAR KEWR 010851Z 27010KT 10SM FEW030 BKN070 BKN100 BKN210 04/M03 A2969 RMK SLP054 T00391033 52012")
```

# Pressure

In our example a pressure value is coded in the **A2972** as inHg (inch of mercury). With the default parameter `altimeter = FALSE`, the function `metar_pressure` returns a pressure in hPa.

```{r, message = TRUE, warning = TRUE}
metar_pressure(my_report)
```

The pressure value can be also presented in a METAR report as Q1008, already in hPa.

```{r, message = TRUE, warning = TRUE}
metar_pressure("EPWA 281830Z 18009KT 140V200 9999 SCT037 03/M01 Q1008 NOSIG")
```

If a pressure is needed in inHg (inch of mercury), the parameter *altimeter* has to be set to TRUE.

```{r, message = TRUE, warning = TRUE}
metar_pressure("EPWA 281830Z 18009KT 140V200 9999 SCT037 03/M01 Q1008 NOSIG", altimeter = TRUE)

```

# Wind shear

Information about wind shear can be extracted with the use of the function `metar_windshear()`

```{r, message = TRUE, warning = TRUE}
metar_windshear("CYWG 172000Z 30015G25KT 3/4SM R36/4000FT/D -SN BLSN BKN008 OVC040 M05/M08 A2992 REFZRA WS RWY36 RMK SF5NS3 SLP134")
```

# Runway visibility

Information about runways visibility can be extracted with the use of the function `metar_rwy_visibility()`, in meters

```{r, message = TRUE, warning = TRUE}
metar_rwy_visibility("CYWG 172000Z 30015G25KT 3/4SM R36/4000FT/D -SN BLSN BKN008 OVC040 M05/M08 A2992 REFZRA WS RWY36 RMK SF5NS3 SLP134")
```

or in feet

```{r, message = TRUE, warning = TRUE}
metar_rwy_visibility("CYWG 172000Z 30015G25KT 3/4SM R36/4000FT/D -SN BLSN BKN008 OVC040 M05/M08 A2992 REFZRA WS RWY36 RMK SF5NS3 SLP134", metric = FALSE)
```

# Putting all together

# Decoding METAR Reports

Let’s come back to our `dm` list with historical METAR reports downloaded at the top of this document. Please notice that in all rows there are dates and hours in the front of METAR reports. It will be parsed and placed in the column `METAR_Date` below.

```{r, message = TRUE, warning = TRUE}
head(dm)
```

Now we can decode all elements and place them in the tibble. It is possible to choose between Metric (the default `metric = TRUE)` or Imperial (`metric = FALSE`) systems. Pressure values can be decoded in hPa (the default `altimeter = FALSE`) or in mmHg (`altimeter = TRUE`).

```{r, message = TRUE, warning = TRUE}
decoded_metars <- metar_decode(dm)
```

The following columns were created:

```{r, message = TRUE, warning = TRUE}
names(decoded_metars)
```

First rows of the tibble with decoded METAR reports:

```{r, message = TRUE, warning = TRUE}
print.data.frame(head(decoded_metars)[,-1])
```